State-space age-structured assessment models provide reliable inferences(?)

### 2022-12-09 Project_0/paper/paper.Rmd
saving work on Project 0 paper

Much is known about the reliabiltiy of state-space models that are linear or Gaussian, but applications in fisheris management are
nonlinear and typically include multiple types of observations with varying distributional assumptions. We know relatively little about
the statistical reliability of such models. Also, there is a wide range of potential random effects structures in assessment models
and we  know little about the ability of information criteria to distinguish among such alternative structures.

But those studies focus primarily on Gaussain 
Reliability of hidden process models

@@

Used the WHAM package commit = XXXXXXX (Stock and Miller 2021, Miller and Stock 2020). This packages has also been used
to configure operating and estimating models for closed loop simulations evaluating index-based asssessment methods (Legault et al. In press).

We completed a simulation study with 24 NAA re operating models, 16 M re operating models, 16 Sel re operating models and 16 q re operating models. 
We simulated N data sets for each operating model.
For each simulated data from each operating model we fit a set of estimating models.

Y estimating models fit to each 
## Operating models

common to all:   
  ages, 
  M
  maturity
  waa
  observation error assumptions
  2.5 Fmsy -> Fmsy after 20 years
  Fmsy througout

@@

NAA_oms
  process error assumtions

Population - 
  Initial population, 
  ages, 
  M
  maturity
  waa

Fishing history
All operating models assumed one of two different fishing histories. 
One : Fishing mortality is equal to Fmsy for the whole 40 year period.
Two : Fishing mortality is 2.5 times Fmsy for the first 20 years then changes to Fmsy for the last 20 years.

Overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenman et al. (20XX). Legault et al. (2023) also used similar approaches to defining fishing mortality histories for operating models.

We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models. All post-recruit productivity components are constant in the NAA and survey catchability process error operating models. Therefore steepness and unfished recruitment are also constant over the time period for those operating models (Miller and Brooks 2021). We specified unfished recruitment = $e^10$ and $F_{\text{MSY}} = F_{40}$ equated to a steepness of 0.68 and a and b parameters X and X?

For operating models with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M.

For operating models with time-varying random effects for fishery selectivity, Fmsy is also not constant however we use the same Fhistory as other operating models which corresposds to Fmsy at the mean selectivity parameters.

Initial population starts at equilibium population structure associated with F= x*Fmsy where x = 1 or 2.5.

surveys, selectivity

fleets, selectivity

M_oms
NOTE: inv_trans_rho function in set_M.R is mis-defined. Will affect correlation parameters assigned in operating models?

@@

32 estimating models
1-20 fit to each NAA RE operating model
5-24 fit to each M RE operating model
5-20,25-28 to each sel RE operating model
5-20, 29-32 to each q RE operating model

@@

\bibliography{sweep-paper}
<!-- \bibliography{sweep-paper}
\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->

>>>>

Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada and the Northeast US \citep{nielsenberg14,cadigan16,stockmiller21}.

Much is known about the reliability of state-space models that are linear or Gaussian \citep{aeberhardetal18}, but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions. We know relatively little about the statistical reliability of such models. Also, there is a wide range of potential random effects structures in assessment models and we  know little about the ability of information criteria to distinguish among such alternative structures.

But those studies focus primarily on Gaussian 
Reliability of hidden process models

@@

Used the WHAM package  \citep[][commit 77bbd94]{stockmiller21} (Miller and Stock 2020). This packages has also been used to configure operating and estimating models for closed loop simulations evaluating index-based assessment methods \citep{legaultetalinpress}.

We completed a simulation study with a number of operating models that can be categorized based on where random effects are assumed: abundance at age, natural mortality, fleet selectivity, or index catchability. For each operating model assumptions about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.

We simulated 100 data sets for each operating model.
For each simulated data from each operating model we fit a set of estimating models.

Y estimating models fit to each

## Operating models

common to all:   
  ages = 1 to 10+, 
  M
  maturity

@@

### Population

There are 10 age classes: ages 1 to 10+.

Spawning was assumed to occur 1/4 of the way through the year.

Natural mortality rate was assumed 0.2 when it was constant and the mean of the time series process for operating models with M random effects.
maturity a50 = 2.89, slope = 0.88

Weight at age was generated with a LVB growth
\[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
\]
where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
\[
W_a = \theta_1 L_a^{\theta_2}
\]
where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$.

We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all operating models. All post-recruit productivity components are constant in the NAA and survey catchability process error operating models. Therefore steepness and unfished recruitment are also constant over the time period for those operating models (Miller and Brooks 2021). We specified unfished recruitment = $R_0 = e^{10}$ and $F_{\text{MSY}} = F_{40} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
\[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\]
Beverton-Holt parameterization.

The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from Wiedenmann et al. (20XX). Legault et al. (2023) also used similar approaches to defining fishing mortality histories for operating models.

Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times F_{\text{MSY}}$ or $F = F_{\text{MSY}}$ for the two alternative fishing histories. That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.

For operating models with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other operating models this equates to a steepness and R0 at the mean of the time series process for M. For operating models with time-varying random effects for fishery selectivity, Fmsy is also not constant however we use the same F history as other operating models which corresponds to Fmsy at the mean selectivity parameters.

### Fleets

We assumed a single fleet operating year round for catch observations with logistic selectivity for the fleet with $a_{50} = 5$ and slope = 1. This selectivity is was used to define $F_{\text{MSY}}$ for the Beverton-Holt stock recruitment parameters above. We assumed a logistic-normal distribution for the age-composition observations for the fleet.

### Fishing histories
All operating models assumed one of two different fishing histories. 
One : Fishing mortality is equal to Fmsy (0.348) for the whole 40 year period.
Two : Fishing mortality is 2.5 times Fmsy for the first 20 years then changes to Fmsy for the last 20 years.

### Indices
Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year). Actually we have it currently configured that both occur 0.5 way through the year. Catchability of both surveys are assumed to be 0.1.  We assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1. We assumed a logistic-normal distribution for the age-composition observations.

### Observation Uncertainty

Standard deviation for log-aggregate catch was 0.1. There were two levels of observation error variance for indices and age composition for both indices and fleet catch. A low uncertainty specification assumed standard deviation of both series of log-aggregate index observations was 0.1 and the standard deviation of the logistic-normal for age composition observations was 0.3 In the high uncertainty specification the standard deviation for log-aggregate indices was 0.4 and that for the age composition observations was 1.5.

### Operating models with random effects on numbers at age

24 operating models, 16 Sel re operating models and 16 q re operating models.
Table of process error assumptions

### Operating models with random effects on natural mortality

16 operating models
Table of process error assumptions

NOTE: inv_trans_rho function in set_M.R is mis-defined. Will affect correlation parameters assigned in operating models?

### Operating models with random effects on fleet selectivity

16 operating models
Table of process error assumptions

### Operating models with random effects on index catchability

16 operating models
Table of process error assumptions

@@

32 estimating models
Table of estimating models

1-20 fit to each NAA RE operating model
5-24 fit to each M RE operating model
5-20,25-28 to each sel RE operating model
5-20, 29-32 to each q RE operating model

@@

\bibliography{paper}
<!-- \bibliography{project-0-paper}
\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->

### 2022-12-09 Project_0/paper/paper.tex
saving work on Project 0 paper

Used the WHAM package commit = XXXXXXX (Stock and Miller 2021, Miller
and Stock 2020). This packages has also been used to configure operating
and estimating models for closed loop simulations evaluating index-based
asssessment methods (Legault et al.~In press).

We completed a simulation study with 24 NAA re operating models, 16 M re
operating models, 16 Sel re operating models and 16 q re operating
models. We simulated N data sets for each operating model. For each
simulated data from each operating model we fit a set of estimating
models.

Y estimating models fit to each \#\# Operating models

common to all:\\
ages, M maturity waa observation error assumptions 2.5 Fmsy
-\textgreater{} Fmsy after 20 years Fmsy througout

@@

NAA\_oms process error assumtions

Population - Initial population, ages, M maturity waa

Fishing history All operating models assumed one of two different
fishing histories. One : Fishing mortality is equal to Fmsy for the
whole 40 year period. Two : Fishing mortality is 2.5 times Fmsy for the
first 20 years then changes to Fmsy for the last 20 years.

Overfishing assumptions is based on average estimates of overfishing for
NE groundfish stocks from Wiedenman et al.~(20XX). Legault et al.~(2023)
also used similar approaches to defining fishing mortality histories for
operating models.

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all operating models. All
post-recruit productivity components are constant in the NAA and survey
catchability process error operating models. Therefore steepness and
unfished recruitment are also constant over the time period for those
operating models (Miller and Brooks 2021). We specified unfished
recruitment = \(e^10\) and \(F_{\text{MSY}} = F_{40}\) equated to a
steepness of 0.68 and a and b parameters X and X?

For operating models with time-varying random effects for M, steepness
is not constant, but we used the same alpha and beta parameters as other
operating models this equates to a steepness and R0 at the mean of the
time series process for M.

For operating models with time-varying random effects for fishery
selectivity, Fmsy is also not constant however we use the same Fhistory
as other operating models which corresposds to Fmsy at the mean
selectivity parameters.

Initial population starts at equilibium population structure associated
with F= x*Fmsy where x = 1 or 2.5.

surveys, selectivity

fleets, selectivity

M\_oms NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will
affect correlation parameters assigned in operating models?

@@

\bibliography{sweep-paper}

>>>>

Used the WHAM package \citep[][commit 77bbd94]{stockmiller21} (Miller
and Stock 2020). This packages has also been used to configure operating
and estimating models for closed loop simulations evaluating index-based
assessment methods \citep{legaultetalinpress}.

We completed a simulation study with a number of operating models that
can be categorized based on where random effects are assumed: abundance
at age, natural mortality, fleet selectivity, or index catchability. For
each operating model assumptions about variance of process errors and
observations are required and the values we used were based on a review
of the range of estimates from recent applications of WHAM in management
of stocks of haddock, butterfish, and American plaice in the NE US.

We simulated 100 data sets for each operating model. For each simulated
data from each operating model we fit a set of estimating models.

Y estimating models fit to each

\hypertarget{operating-models}{%
\subsection{Operating models}\label{operating-models}}

common to all:\\
ages = 1 to 10+, M maturity

@@

\hypertarget{population}{%
\subsubsection{Population}\label{population}}

There are 10 age classes: ages 1 to 10+.

Spawning was assumed to occur 1/4 of the way through the year.

Natural mortality rate was assumed 0.2 when it was constant and the mean
of the time series process for operating models with M random effects.
maturity a50 = 2.89, slope = 0.88

Weight at age was generated with a LVB growth \[
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
\] where \(t_0 = 0\), \(L_\infty = 85\), and \(k = 0.3\), and a L-W
relationship such that \[
W_a = \theta_1 L_a^{\theta_2}
\] where \(\theta_1 = e^{-12.1}\) and \(\theta_2 = 3.2\).

We assumed a Beverton-Holt stock recruit function with constant
pre-recruit mortality parameters for all operating models. All
post-recruit productivity components are constant in the NAA and survey
catchability process error operating models. Therefore steepness and
unfished recruitment are also constant over the time period for those
operating models (Miller and Brooks 2021). We specified unfished
recruitment = \(R_0 = e^{10}\) and \(F_{\text{MSY}} = F_{40} = 0.348\)
equated to a steepness of 0.69 and \(\alpha=0.60\) and
\(\beta = 2.4 \times 10^{-5}\) for the \[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\] Beverton-Holt parameterization.

The magnitude of the overfishing assumptions is based on average
estimates of overfishing for NE groundfish stocks from Wiedenmann et
al.~(20XX). Legault et al.~(2023) also used similar approaches to
defining fishing mortality histories for operating models.

Initial population was configured at the equilibrium distribution
fishing at either \(F = 2.5\times F_{\text{MSY}}\) or
\(F = F_{\text{MSY}}\) for the two alternative fishing histories. That
is for a deterministic model, the age composition would not change over
time when the fishing mortality was constant at the respective level.

For operating models with time-varying random effects for M, steepness
is not constant, but we used the same alpha and beta parameters as other
operating models this equates to a steepness and R0 at the mean of the
time series process for M. For operating models with time-varying random
effects for fishery selectivity, Fmsy is also not constant however we
use the same F history as other operating models which corresponds to
Fmsy at the mean selectivity parameters.

\hypertarget{fleets}{%
\subsubsection{Fleets}\label{fleets}}

We assumed a single fleet operating year round for catch observations
with logistic selectivity for the fleet with \(a_{50} = 5\) and slope =
1. This selectivity is was used to define \(F_{\text{MSY}}\) for the
Beverton-Holt stock recruitment parameters above. We assumed a
logistic-normal distribution for the age-composition observations for
the fleet.

\hypertarget{fishing-histories}{%
\subsubsection{Fishing histories}\label{fishing-histories}}

All operating models assumed one of two different fishing histories. One
: Fishing mortality is equal to Fmsy (0.348) for the whole 40 year
period. Two : Fishing mortality is 2.5 times Fmsy for the first 20 years
then changes to Fmsy for the last 20 years.

\hypertarget{indices}{%
\subsubsection{Indices}\label{indices}}

Two time series of surveys are assumed and observed in numbers rather
than biomass for the entire 40 year period with one occurring in the
spring (0.25 way through the year) and one in the fall (0.75 way through
the year). Actually we have it currently configured that both occur 0.5
way through the year. Catchability of both surveys are assumed to be
0.1. We assumed logistic selectivity for both indices with
\(a_{50} = 5\) and slope = 1. We assumed a logistic-normal distribution
for the age-composition observations.

\hypertarget{observation-uncertainty}{%
\subsubsection{Observation Uncertainty}\label{observation-uncertainty}}

Standard deviation for log-aggregate catch was 0.1. There were two
levels of observation error variance for indices and age composition for
both indices and fleet catch. A low uncertainty specification assumed
standard deviation of both series of log-aggregate index observations
was 0.1 and the standard deviation of the logistic-normal for age
composition observations was 0.3 In the high uncertainty specification
the standard deviation for log-aggregate indices was 0.4 and that for
the age composition observations was 1.5.

\hypertarget{operating-models-with-random-effects-on-numbers-at-age}{%
\subsubsection{Operating models with random effects on numbers at
age}\label{operating-models-with-random-effects-on-numbers-at-age}}

24 operating models, 16 Sel re operating models and 16 q re operating
models. Table of process error assumptions

\hypertarget{operating-models-with-random-effects-on-natural-mortality}{%
\subsubsection{Operating models with random effects on natural
mortality}\label{operating-models-with-random-effects-on-natural-mortality}}

16 operating models Table of process error assumptions

NOTE: inv\_trans\_rho function in set\_M.R is mis-defined. Will affect
correlation parameters assigned in operating models?

\hypertarget{operating-models-with-random-effects-on-fleet-selectivity}{%
\subsubsection{Operating models with random effects on fleet
selectivity}\label{operating-models-with-random-effects-on-fleet-selectivity}}

16 operating models Table of process error assumptions

\hypertarget{operating-models-with-random-effects-on-index-catchability}{%
\subsubsection{Operating models with random effects on index
catchability}\label{operating-models-with-random-effects-on-index-catchability}}

16 operating models Table of process error assumptions

@@

\bibliography{paper}
